Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-04-21 Manuscript/manuscript.Rmd
editted manuscript
manuscript edits

Replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1). The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained.

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix (Figure 1). For example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. For example, for 10 kb data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

### FEATURE ENGINEERING AND RE-SAMPLING

@@

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance in bases (refered to hereafter simply as $\textit{distance}$), from the center of the bin to the center of the nearest region defined by a specific genomic annotation was calculated. The log transformation was used as a normalization technique in order to account for the skewness of the data (Supplementary Figures 1).

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were performed to fully compare re-sampling techniques. To reduce bias due to random dataset generation, 10-fold cross validation was used <!--to tune the number of features to consider at each node and the number of ensemble trees to aggregate-->.The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was assessed using the test set, and performance metrics were recorded.

The efficacy of each re-sampling technique was compared using composite metrics including F1-score and MCC. These metrics are defined as follows:

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. <!--Additional threshold-free metrics where also reported including the AUROC and AUPRC.--> Lastly, for the task of identifying the genomic annotations that were most predictive of the formation of TAD boundaries, we ranked the variable importances associated with each feature of the ensemble system-specific random forest model. <!--To calculate variable importances for each model, the F1-Score on the out-of-bag portion of the data within each tree was recorded. Then the same was done after permuting each predictor variable. The difference between the two AUPRC values was then averaged over all trees, and normalized by the standard error of the differences.--> Annotations with low rankings correspond to functional epigenomic features that are either positively or negatively enriched near TAD boundaries, and thus, better allow the model to discriminate between genomic bins that overlap with flanked TAD boundaries.

@@

We examined the class imbalance of the different data resolutions for each cell line. As expected, the total number of genomic bins decreased as resolution decreased (Figure 4A). Specifically, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins. The imbalance ratios, however, remained relatively stable and consistent across resolutions and cell line (Figure 4B). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Here, the low imbalance ratios indicated a wide disparity between the number of genomic bins that overlap with a flanked TAD boundary and the much larger number that did not.

@@

We first evaluated performances for models implemented using OC predictor types, without re-sampling, as these characteristics represented models with the least granularity and could act as a baseline. In evaluating such composite metrics like F1-Score and MCC, we found that these models exhibited poor performances across resolutions, for the GM12878 cell line (Figure 5A). <!--The average F1-Score and MCC across resolution was found to be 0.04 and 0.07 respectively.--> When including OP perdictor types as the feature space, coupled with no resampling, we found that performances were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). <!--Here, the average F1-Score and MCC was found to be 0.01 and 0.03 respectively.--> Models built on distance-type predictors in conjuction with no resampling, however, were found to exhibit slightly better performances than the other two types<!--, with average F1-Scores and MCCs of 0.08 and 0.10 respectively--> (Figure 5C). Importantly, when validating models built on fully imbalanced datasets, the F1-Score and MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Furthermore, evaluating models using conventional metrics such as AUC or accuracy did not depict the true severity in performance (Supplementary Table 3). <!--The average AUC across resolutions for the OC, OP, and distance type predictors was 0.65, 0.66, and 0.85 respectively, where as the average accuracies were 0.89, 0.89, and 0.90 respectively (Supplementary Table 3).--> Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were unable to be reported due to the models inability to produce any true positive cases.

@@

Distance profiles on a $log_{2}$ scale were found to exhibit significantly different densities between classes of genomic bins for TFBS with highly ranked predictive importances. Using permutation tests we found that the transcription factors SMC3a, ZNF143, CTCF, and RAD21 tended to have binding peaks that that were located closer to the centers of 10 kb sized genomic bins that overlapped with flanked TAD boundaries than to bins that did not overlap with flanked TAD boundaries across both cell lines (Figure 7 A-H). Similar distance profiles for the same TFBS were obtained when assessed at lower resolutions, across both cell lines (Supplementary Figures X-X)

>>>>

Replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1). The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. For each of the 4 data resolutions, a total of 8008, 4869, 2449, and 1075 unique TADs were extracted from the GM12878 cell line, and a total of 5828, 4097, 2181, and 986 unique TADs were extracted from the K562 cell line. The TADs were distributed throughput the autosomal chromosomes, with larger chromosomes containing more TADs (Supplementary Figure 14; Supplementary Table 8). The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained.

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix (Figure 1). For example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. For example, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0). The process was repeat for each subsequent resolution.

@@

### FEATURE ENGINEERING

@@

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance in bases (refered to hereafter simply as $\textit{distance}$), from the center of the bin to the center of the nearest region defined by a specific genomic annotation was calculated. The log transformation was used as a normalization technique in order to account for the skewness present in the distance densities (Supplementary Figures 1).

### DATA RE-SAMPLING

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were performed to fully compare re-sampling techniques. To reduce bias due to random dataset generation, 10-fold cross validation was used <!--to tune the number of features to consider at each node and the number of ensemble trees to aggregate-->. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was assessed using the test set, and performance metrics were recorded.

The efficacy of each ensemble system in the framework was compared using composite metrics including F1-score and MCC. These metrics are defined as follows:

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. Additional performance metrics were reported in the supplementary materials.

<!--Additional threshold-free metrics where also reported including the AUROC and AUPRC.-->

<!--
Lastly, for the task of identifying the genomic annotations that were most predictive of the formation of TAD boundaries, we ranked the variable importances associated with each feature of the ensemble system-specific random forest model. To calculate variable importances for each model, the F1-Score on the out-of-bag portion of the data within each tree was recorded. Then the same was done after permuting each predictor variable. The difference between the two AUPRC values was then averaged over all trees, and normalized by the standard error of the differences. Annotations with low rankings correspond to functional epigenomic features that are either positively or negatively enriched near TAD boundaries, and thus, better allow the model to discriminate between genomic bins that overlap with flanked TAD boundaries.
-->

@@

To assess the full severity of class imbalance present when constructing the models, we investigated the relationship between the total number of bins (sample observations) included at each resolution and the subsequent imbalance ratio (IR). As expected, the total number of genomic bins decreased as resolution decreased (Figure 4A). Specifically, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins. however, the IRs remained relatively stable and consistent across resolutions and cell lines (Figure 4B). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Here, the low imbalance ratios indicated a wide disparity between the number of genomic bins that overlap with a flanked TAD boundary and the much larger number that did not.

@@

In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. In evaluating such composite metrics like F1-Score and MCC, we found that for the GM12878 cell line, these models exhibited poor performances across all resolutions (Figure 5A). <!--The average F1-Score and MCC across resolution was found to be 0.04 and 0.07 respectively.-->

We then gradually increased the granularity in the characteristics of the model by assess the model built on OP predictor types, while still not performing any data resampling. Here we found that performances were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). <!--Here, the average F1-Score and MCC was found to be 0.01 and 0.03 respectively.--> Finally, models built on distance-type predictors in conjuction with no resampling, however, were found to exhibit slightly better performances than the other two types<!--, with average F1-Scores and MCCs of 0.08 and 0.10 respectively--> (Figure 5C).

Importantly, when validating these 3 models built on fully imbalanced datasets, the F1-Score and MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Furthermore, we found that evaluating models using conventional metrics, such as AUC or accuracy, did not depict the true severity in performance (Supplementary Table 3). <!--The average AUC across resolutions for the OC, OP, and distance type predictors was 0.65, 0.66, and 0.85 respectively, where as the average accuracies were 0.89, 0.89, and 0.90 respectively (Supplementary Table 3).--> Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were also unable to be reported due to the models inability to produce any true positive cases.

@@

Distance profiles on a $log_{2}$ scale were found to exhibit significantly different densities between classes of genomic bins for TFBS with highly ranked predictive importances. Using permutation tests we found that the transcription factors SMC3a, ZNF143, CTCF, and RAD21 tended to have binding peaks that that were located closer to the centers of 10 kb sized genomic bins that overlapped with flanked TAD boundaries than to bins that did not overlap with flanked TAD boundaries across both cell lines (Figure 7 A-H). Similar distance profiles for the same TFBS were obtained when assessed at lower resolutions, across both cell lines (Supplementary Figures 12 & 13 A-L).
